Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-01-12 Manuscript_Nater_etal.qmd
edits to qdm and bib

Quantifying the temporal and spatial variation in animal demography is a central theme of ecology and important for directing policy. However, field sampling at large spatial extents for long periods of time is cost prohibitive and often politically untennable. Citizen science programs can alleviate these issues by harnessing stakeholders to increase the spatial and temporal resolution of datasets. Yet, many citizen science programs lack data on individual animals making it difficult to model demography. Integrated population models (IPM) offer promise to use multiple streams of data on populations or multiple aspects of a single dataset to estimate key vital rates for populations or communities. We leveraged a large-scale citizen science monitoring program---the "Hønsefuglportalen"--- to estimate spatial and temporal variation in demography for willow ptarmigan (*Lagopus lagopus*). The core field sampling approach was line-transect distance sampling with age-specific detections---the latter critical to modelling the population dynamics in the IPM. We explored latitudinal and longitudinal variation in recruitment, survival, and density; patterns of density dependence; and the relationship between recruitment and rodent occupancy using an IPM in a Bayesian framework following open science principles. Overall results were consistent with expectations of ecological theory. Slower life history patterns were observed at higher latitudes and altitudes portending differential effects of climate change on ptarmigan across their range. Observed patterns between rodents and recruitment were consistent with the alternative prey hypothesis (APH); however, spatial variation in the directionality and magnitude of effect were substantial. The robustness of the hierarchical structure of our model was instrumental in elucidating these patterns. Our study highlights the power of citizen science, IPMs, and open science principles to estimate spatial and temporal variation in the demography for a culturally important species.

@@

Building on the work of @nilsen2023, we applied a novel integrated population model to data collected through a national-scale participatory monitoring programme to estimate spatial and temporal variation in demography of a culturally important game bird species, the willow ptarmigan. While our study was exploratory in nature, it recovered patterns consistent with ecological and life-history theory including trade-offs between survival and recruitment, and a tendency towards slower life histories at higher latitudes and altitudes. Space-for-time substitution also provided the statistical power necessary for the analysis to provide evidence for the alternative prey hypothesis, i.e. ptarmigans benefiting from high abundance of alternative prey for their predators. <!-- To me, this little paragraph has "abstract vibes", or maybe "conclusion vibes". I'll leave it here for now but note that it may get moved eventually. 
<!-- Matter of style but this is typically how I do it but your call. :-) >
## Abundance and vital rates across space and time

@@

Density dependence is an important process to understand in ecology but more so for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; aanes2002; @willebrand2001]. Negative density dependence has been found in numerous gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]).

While we found the highest population growth rates in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A], our analysis did not provide evidence for a strong relationship between density and growth rates across space. However, this is unsurprising as we did find substantial spatial differences in demography, which suggests that carrying capacity too varies over space. Also, support for density dependence in willow ptarmigan has been mixed. @myrberget1988 observed no change in productivity despite a 50% decrease in abundance. Conversely, @pedersen2004 found strong overwinter negative density dependence and they posited that immigration was the vital rate that responded to changes in density. Low survival rates appeared to have a one-year lag following abundance peaks. This pattern could be a result of negative density dependence that went undetected when looking at the ensemble of samples across the 40 areas. Negative density dependence has been shown in willow ptarmigan populations which may explain explain reduced survival following years with elevated densities [@pedersen2004].

<!--# Not happy with the DD para yet but I can work on Jan 11 -->
<!--# @JAMES: Feel free to add more, but I feel like if we want to go for a DD deep-dive, I need to run a test for evidence of WITHIN area DD first. Because there may well be a pattern there... -->

<!-- Maybe another sentence or two about the low survival following the peak, which could be either due to negative density-dependence leading to resource competition among ptarmigan, or due to the rodent peak allowing high recruitment also for predators, which then in turn will have increased impacts on all prey populations in the subsequent year? -->

@@

@@

Population densities in all areas were subject to substantial variation across years, featuring periods of stability, increase, and decrease (SF "TimeSeries_PopDens1.pdf"). In most areas, there were also strong year-by-year fluctuations in population density on top of longer-term trends. Some of the resulting "high density years" were highly synchronized across large spatial scales, such as the years 2011, 2013, and 2018. Taking a closer look, we find that these are years that are characterized by high recruitment (SF "TimeSeries_rRep.pdf") and low subsequent survival (SF "TimeSeries_pSurv.pdf"), which often resulted in steep population declines towards the following year. The fact that these same years also match up with observed peaks in rodent abundance in many areas provides evidence for the Alternative Prey Hypothesis (APH). The AHP stipulates that high abundance of alternative prey (rodents, in this case) for common predators leads to population growth, and is well-supported throughout the literature for a range of taxa [e.g., @hagen1952; @Kjellander_Nordström_2003; @Reif_Tornberg_Jungell_Korpimäki_2001], including willow ptarmigan [@bowler2020]. Conversely, @Nyström_etal_2006 suggested that gyrfalcons (a specialized ptarmigan predator) did not respond to rodent populations and did not switch to alternative prey when ptarmigan populations are low. Nonetheless, other predators such as fox could prey switch in response to changing availability of rodents and ptarmigan as shown by @Breisjøberget_etal_2018 and @bowler2020. However, statistical associations in our study and others is not causal and alternative explanations for this pattern are plausible. The positive association between rodents and ptarmigan could be caused by bottom-up processes such as food availability unrelated to predation. Future work should develop causal models and manipulative experiments to establish causal relationships among top-down and bottom-up processes.

Taken together, this alludes to the large role small rodents play in shaping ptarmigan demography and population dynamics and is further supported by the largely positive effects of rodent abundance on recruitment estimated by our model (@fig-paramCorrs C). Once more, the highest latitude and highest altitude areas stood out, sporting the strongest rodent effects (SF "betaR_Map.pdf"). This may be linked to the overall low ptarmigan densities in these areas, as e.g. @Nielsen_1999 found evidence that gyrfalcon predation on ptarmigan was most impactful when ptarmigan densities were low. <!-- This previous sentence is adapted from a suggestion by James, but I feel like we need some more explanation/arguments here to clarify this. Perhaps also some other and/or more specific hypotheses= --> Notably, the conclusive estimation of overall positive effects of rodent abundance on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas. When @nilsen2023 fit the IDSM to data from only a single area, they were unable to obtain a reliable estimate for the rodent effect due to limited statistical power. Consequently, the space-for-time substitution that comes with extending the model across multiple area allows estimation of covariate effects that otherwise cannot be estimated, and opens up for future possibilities for studying effects of not just rodents, but also other environmental drivers on ptarmigan population dynamics. For example, \[...\] <!-- @ERLEND here we could add some thoughts on other covariates that may matter and are realistic/relevant/realistic to include in future studies. Best if it kind of naturally leads over to (and does not overlap too much with) the next paragraph  -->

@@

Extending the present analysis by including additional covariates may also help with better understanding the mechanisms underlying the large portion of demographic rate variation that could only be attributed to random variation so far. This is the case especially for the relatively large residual variation [@fig-varDecomp] but also relevant for constant spatial and shared temporal variation. Interestingly, we found that a larger portion of variation in recruitment was allocated to spatial than to temporal factors, while the opposite was the case for survival. \[...\] <!-- Any hypotheses/thoughts related to large- vs. local-scale drivers to explain this? -->

@@

@@

Management decisions made at the resolution of large geo-political boundaries (e.g., Norway) run a high risk of being inadequate when there is substantial spatial variation in demographic processes and population dynamics, as is the case for willow ptarmigan. In Norway, willow ptarmigan -- and small game in general -- is managed \[...\] <!-- Brief summary of how and at what spatial scales patrmigan/small game management works in Norway --> The results from ur study suggests that management decisions should be spatially refined and ideally informed by knowledge about the "local" population processes. \[...\] <!-- Some specific implications of the results for management of hunting and of habitat (e.g. better managed hunting is good because we do see increases in majority of local populations, but the mountain areas have had recent declines that may call for additional local measures, these measures may need to be adjusted since the mountain populations appear to have slower life histories and likely different sensitivities, etc.) -->

Besides pointers on how to adjust management of ptarmigan and their habitat, our results also provided some insights into the value, and possibly opportunities for improving the monitoring programme. First and foremost, our study demonstrates the tremendous potential lying in coordinating structured monitoring using a common protocol and training programme, as well as data processing pipeline. These were indeed the prerequisites that allowed us to easily and efficiently integrate data collected across the entire country in a joint analysis, and draw inference on spatio-temporal variation in demography and population dynamics at a large scale but also with good spatial resolution. While overall less variable across space and time than vital rates, differences in detection probabilities were nonetheless evident (SFs "Avg_detect_Map.pdf" and "TimeSeries_pDetect.pdf") and may help with mapping out potential for improvement in the monitoring programme. Particularly, we found generally lower detection probabilities in the northern half of Norway. \[...\] <!-- Some thoughts on why that might be. Could be related to landscape features and/or available observer pool. Any suggestions on what could be done to improve detection in the North, or at least "make up" for it? --> Ultimately, increased detection probability would contribute to obtaining more precise estimates of both population density and demographic rates, which -- in turn -- would be of great value in particular in areas with relatively low population densities, low number of transects, and less years of data.

@@

The primary focus of this work was placed on developing an effective pipeline for integrating data and modelling population dynamics across a large number of areas. Consequently, many additional opportunities for improving and refining the modelling framework itself remain. 
First, the precision and accuracy of model estimates might be increased through better accounting for heterogeneity and potential biases in detection of birds during the line transect surveys. In an earlier study analyzing data from the same monitoring programme, @bowler2020, for example, found that detection probability was not independent of the size of group birds were part of, resulting in birds in larger groups being more likely to be detected, especially at larger distances. When birds are observed in larger groups, it is also not unlikely that human observers may miscount, i.e. that there is some observation error in the number reported. This could be accounted for by adding an additional layer of hierarchy to observation process (see e.g. REF), which could further be used to also account for error in judging the observation distance [e.g. @marques2004]. Another potential source of bias in our IDSM is related to failure to correctly assign the age class of observed birds. @nilsen2023 showed that incorrect age assessment can bias (relative) estimates of survival and recruitment, and while they only found a weak bias in their case study on a single area, the problem may be larger in a multi-area setting that may contain areas with different proportions of misclassified observations. If misclassification happened at random, mixture models could be used to determine the likely age class of individuals to whom no age class was assigned during observations (+ref). In our case, however, juveniles are likely more prone to misclassification due to [...] <!-- @ERLEND can you add reasons? -->. Future studies should investigate to what degree available information on e.g. group composition could be used for this, and what kind of auxiliary data would need to be collected to reliably model misclassification error.

The second (and perhaps most attractive) aspect of our modelling framework in the context of future work is its spatio-temporal hierarchical structure. While we included spatial, temporal, and residual variation in our framework here, we treated them as independent. Alternatively, spatial (and temporal) correlations among parameters can be modelled explicitly, something that is commonly done e.g. for modern species distribution models [e.g. @pacifici2017; @guelat2018]. For demographic models, this has rarely been implemented so far, not least due to the fact that few demographic models have sufficient spatial resolution [@schaub2021]. The ptarmigan IDSM presented in this study, however, does have sufficient resolution and our results do indeed support that there is spatial clustering in both overall and time-dependent demographic parameters (e.g. @fig-pVR, SFs "Avg_pSurv_Map.pdf" & "Avg_rRep_Map.pdf"). Furthermore, we did find that mixing of several of the global and area-specific intercept parameters in the current model was suboptimal, suggesting that there may be much to gain from additional structuring, as well as from development of more efficient MCMC sampling strategies for the resulting extended model. One promising framework for approaching this are conditionally autoregressive models (CARs). Such models have been used repeatedly for modelling spatial autocorrelation in species occupancy and demographic rates [e.g. @saracco2010; @saracco2012; @guelat2018]
and are straightforward to implement using NIMBLE [@lawson2020]. One possible challenge with using CAR models to explicitly model spatial correlations within our ptarmigan IDSM is that CAR models rely on "neighborhood" relationships between discrete areas and many "neighbors" are missing in our ptarmigan data (e.g. @fig-dens). While estimation of latent parameters in missing areas may possible (Perry de Valpine, personal communication; @schaub2021 chapter 19), this also provides an opportunity for inclusion of additional data. The line transect survey data included in this study constitutes just the publicly available part of the data collected through "Hønsefuglportalen" but the programme also includes additional surveys on private land. [...] <!-- @ERLEND: Can we add a very brief ocverview over what the coverage (in terms of municipalities) would be if we had access to all the Hønsefuglportalen data? --> Exploring to what degree additional data from Hønsefuglportalen could be included in future studies employing an extended IDSM with additional spatial structuring is therefore a worthwhile endeavor.

Finally, including further data beyond the line transect surveys may be relevant in the future, and in particular in the context of informing and improving management of ptarmigan hunting. 
In the present study, we have used auxiliary radio-telemetry data to supplement information on survival, but since this data was available for only one out of 41 areas, its influence was likely small. Nonetheless, this illustrates a way for how smaller datasets from single or subsets of areas can be integrated into a large-scale modelling framework. Other relevant data could be included using the same approach [...]  <!-- @ERLEND: Any thoughts on what the status here is? Might there be other data available from Lierne and/or beyond (other telemetry data, nest monitoring, etc.? Are there exciting future prospects e.g. linked to automated monitoring with cameras/sound boxes at a larger scale? -->
The most relevant source of data to be included into the IDSM framework in the near future, however, is harvest data. <!-- @ERLEND 1-2 sentence summary of what kind of harvest data is available, and at what spatial covarage and resolution --> As the IDSM framework is, in essence, an IPM, harvest can be modelled through partitioning of survival into cause-specific mortality in the process model and inclusion of relevant harvest data likelihoods [e.g. @gamelon2021; @nater2021]. Harvest effects on willow ptarmigan have been studied previously yet uncertainties persist [@sandercock_etal2011; aanes2002; @pedersen2004; @willebrand2001]. [...] <!-- @JAMES & Erlend: Maybe we can be a bit more specific about what the "uncertainties" are regarding ptarmigan harvest? Might be good to keep the "spatially explicit" perspective a bit, i.e. why would it be great to look at this with this kind of model, instead of just aggregate numbers across large regions? Also, maybe we should highlight that up-to-date (easilty repeatable) analyses are important because things can change, and harvest effects may interact with other (emergent) factors such as climate change and habitat degradation. Finally, it could also be nice to link this to some studies on other species... -->

@@

>>>>

Quantifying the temporal and spatial variation in animal demography is a central theme of ecology and important for directing policy. However, field sampling at large spatial extents for long periods of time is cost prohibitive and often politically untenable. Citizen science programs can alleviate these issues by harnessing stakeholders to increase the spatial and temporal resolution of datasets. Yet, many citizen science programs lack data on individual animals making it difficult to model demography. Integrated population models (IPM) offer promise to use multiple streams of data on populations or multiple aspects of a single dataset to estimate key vital rates for populations or communities. We leveraged a large-scale citizen science monitoring program---the "Hønsefuglportalen"--- to estimate spatial and temporal variation in demography for willow ptarmigan (*Lagopus lagopus*). The core field sampling approach was line-transect distance sampling with age-specific detections---the latter critical to modelling the population dynamics in the IPM. We explored latitudinal and longitudinal variation in recruitment, survival, and density; patterns of density dependence; and the relationship between recruitment and rodent occupancy using an IPM in a Bayesian framework following open science principles. Overall results were consistent with expectations of ecological theory. Slower life history patterns were observed at higher latitudes and altitudes portending differential effects of climate change on ptarmigan across their range. Observed patterns between rodents and recruitment were consistent with the alternative prey hypothesis (APH); however, spatial variation in the directionality and magnitude of effect were substantial. The robustness of the hierarchical structure of our model was instrumental in elucidating these patterns. Our study highlights the power of citizen science, IPMs, and open science principles to estimate spatial and temporal variation in the demography for a culturally important species.

@@

## Abundance and vital rates across space and time

@@

Density dependence is an important process to understand in ecology but more so for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @aanes2002; @willebrand2001]. Negative density dependence has been found in numerous gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]). While we found the highest population growth rates in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A], our analysis did not provide evidence for a strong relationship between density and growth rates across space. However, this is unsurprising as we did find substantial spatial differences in demography, which suggests that carrying capacity too varies over space. Also, support for density dependence in willow ptarmigan has been mixed. @myrberget1988 observed no change in productivity despite a 50% decrease in abundance. Conversely, @pedersen2004 found strong overwinter negative density dependence and they posited that immigration was the vital rate that responded to changes in density. Low survival rates appeared to have a one-year lag following abundance peaks. This pattern could be a result of negative density dependence that went undetected when looking at the ensemble of samples across the 41 areas. Future work should explore different time-lags for density and population growth and the possibility for non-linear effects.

<!--# @JAMES: Feel free to add more, but I feel like if we want to go for a DD deep-dive, I need to run a test for evidence of WITHIN area DD first. Because there may well be a pattern there... -->

<!--# Yeah I agree. We already have a lot going on so I am ambivalent about adding another aspect. -->

@@

<!--# Yeah I wasn't able to find much. -->

@@

Population densities in all areas were subject to substantial variation across years, featuring periods of stability, increase, and decrease (SF "TimeSeries_PopDens1.pdf"). In most areas, there were also strong year-by-year fluctuations in population density on top of longer-term trends. Some of the resulting "high density years" were highly synchronized across large spatial scales, such as the years 2011, 2013, and 2018. Taking a closer look, we find that these are years that are characterized by high recruitment (SF "TimeSeries_rRep.pdf") and low subsequent survival (SF "TimeSeries_pSurv.pdf"), which often resulted in steep population declines towards the following year. The fact that these same years also match up with observed peaks in rodent abundance in many areas provides evidence for the Alternative Prey Hypothesis (APH; [@Israelsen2020]). The AHP stipulates that high abundance of alternative prey (rodents, in this case) for common predators leads to population growth, and is well-supported throughout the literature for a range of taxa [e.g., @hagen1952; @Kjellander_Nordström_2003; @Reif_Tornberg_Jungell_Korpimäki_2001], including willow ptarmigan [@bowler2020]. Conversely, @Nyström_etal_2006 suggested that gyrfalcons (a specialized ptarmigan predator) did not respond to rodent populations and did not switch to alternative prey when ptarmigan populations are low. Nonetheless, other predators such as fox could prey switch in response to changing availability of rodents and ptarmigan as shown by @Breisjøberget_etal_2018 and @bowler2020. However, statistical associations in our study and others is not causal and alternative explanations for this pattern are plausible. The positive association between rodents and ptarmigan could be caused by bottom-up processes such as food availability unrelated to predation. Future work should develop causal models and manipulative experiments to establish causal relationships among top-down and bottom-up processes.

Taken together, this alludes to the large role small rodents play in shaping ptarmigan demography and population dynamics and is further supported by the largely positive effects of rodent abundance on recruitment estimated by our model (@fig-paramCorrs C). Once more, the highest latitude and highest altitude areas stood out, sporting the strongest rodent effects (SF "betaR_Map.pdf"). This may be linked to the overall low ptarmigan densities in these areas, as e.g. @Nielsen_1999 found evidence that gyrfalcon predation on ptarmigan was most impactful when ptarmigan densities were low. <!-- This previous sentence is adapted from a suggestion by James, but I feel like we need some more explanation/arguments here to clarify this. Perhaps also some other and/or more specific hypotheses= --> Notably, the conclusive estimation of overall positive effects of rodent abundance on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas. When @nilsen2023 fit the IDSM to data from only a single area, they were unable to obtain a reliable estimate for the rodent effect due to limited statistical power. Consequently, the space-for-time substitution that comes with extending the model across multiple area allows estimation of covariate effects that otherwise cannot be estimated, and opens up for future possibilities for studying effects of not just rodents, but also other environmental drivers on ptarmigan population dynamics. For example, the cover of ericaceous shrubs as a proxy for food availability or spatially-explicit spring green up dates both measured using remote sensing are ecologicaly plausible and feasibly acquired factors. <!-- @ERLEND here we could add some thoughts on other covariates that may matter and are realistic/relevant/realistic to include in future studies. Best if it kind of naturally leads over to (and does not overlap too much with) the next paragraph  -->

<!--# I added two that came to my mind  -->

@@

Extending the present analysis by including additional covariates may also help with better understanding the mechanisms underlying the large portion of demographic rate variation that could only be attributed to random variation so far. This is the case especially for the relatively large residual variation [@fig-varDecomp] but also relevant for constant spatial and shared temporal variation. Interestingly, we found that a larger portion of variation in recruitment was allocated to spatial than to temporal factors, while the opposite was the case for survival. Contrastingly, ptarmigan nest failure is almost entirely caused by predation which has been shown to vary greatly across years and less across space [@Munkebye_Pedersen_Steen_Brøseth_2003; @kvasnes2015; @kvasnes2014] whereas adult survival has been shown to not very greatly across years [@Israelsen2020]. Also, immigration, included in our recruitment term <!--# this true? -->, has been hypothesized to be a spatial process [@kvasnes2015]. This conundrum could be a result of the limited spatial replication of many previous studies that we were able to address with our approach. Future research could help rectify these discrepancies by including more mechanistic relationships between environmental factors and vital rates.

<!-- Any hypotheses/thoughts related to large- vs. local-scale drivers to explain this? -->

@@

<!-- I don't know if I helped or not. It is an interesting result. Erlend, you got thoughts? -->

@@

Management decisions made at the resolution of large geo-political boundaries (e.g., Norway) run a high risk of being inadequate when there is substantial spatial variation in demographic processes and population dynamics, as is the case for willow ptarmigan. In Norway, willow ptarmigan -- and small game in general -- is managed \[...\] <!-- Brief summary of how and at what spatial scales patrmigan/small game management works in Norway --> The results from our study suggests that management decisions should be spatially refined and ideally informed by knowledge about the "local" population processes. \[...\] <!-- Some specific implications of the results for management of hunting and of habitat (e.g. better managed hunting is good because we do see increases in majority of local populations, but the mountain areas have had recent declines that may call for additional local measures, these measures may need to be adjusted since the mountain populations appear to have slower life histories and likely different sensitivities, etc.) -->

Besides pointers on how to adjust management of ptarmigan and their habitat, our results also provided some insights into the value, and possibly opportunities for improving the monitoring programme. First and foremost, our study demonstrates the tremendous potential lying in coordinating structured monitoring using a common protocol and training programme, as well as data processing pipeline. These were indeed the prerequisites that allowed us to easily and efficiently integrate data collected across the entire country in a joint analysis, and draw inference on spatio-temporal variation in demography and population dynamics at a large scale but also with good spatial resolution. While overall less variable across space and time than vital rates, differences in detection probabilities were nonetheless evident (SFs "Avg_detect_Map.pdf" and "TimeSeries_pDetect.pdf") and may help with mapping out potential for improvement in the monitoring programme. Particularly, we found generally lower detection probabilities in the northern half of Norway. \[...\] <!-- Some thoughts on why that might be. Could be related to landscape features and/or available observer pool. Any suggestions on what could be done to improve detection in the North, or at least "make up" for it? -->

<!--# They obviously use a lesser breed of bird dog up that way.Probably use more german shorthairs or vislas. :-)  -->

Ultimately, increased detection probability would contribute to obtaining more precise estimates of both population density and demographic rates, which -- in turn -- would be of great value in particular in areas with relatively low population densities, low number of transects, and less years of data.

@@

The primary focus of this work was placed on developing an effective pipeline for integrating data and modelling population dynamics across a large number of areas. Consequently, many additional opportunities for improving and refining the modelling framework itself remain. First, the precision and accuracy of model estimates might be increased through better accounting for heterogeneity and potential biases in detection of birds during the line transect surveys. In an earlier study analyzing data from the same monitoring programme, @bowler2020, for example, found that detection probability was not independent of the size of group birds were part of, resulting in birds in larger groups being more likely to be detected, especially at larger distances. When birds are observed in larger groups, it is also not unlikely that human observers may miscount, i.e. that there is some observation error in the number reported. This could be accounted for by adding an additional layer of hierarchy to observation process (see e.g. REF), which could further be used to also account for error in judging the observation distance [e.g. @marques2004]. Another potential source of bias in our IDSM is related to failure to correctly assign the age class of observed birds. @nilsen2023 showed that incorrect age assessment can bias (relative) estimates of survival and recruitment, and while they only found a weak bias in their case study on a single area, the problem may be larger in a multi-area setting that may contain areas with different proportions of misclassified observations. If misclassification happened at random, mixture models could be used to determine the likely age class of individuals to whom no age class was assigned during observations (+ref). In our case, however, juveniles are likely more prone to misclassification due to \[...\] <!-- @ERLEND can you add reasons? -->. Future studies should investigate to what degree available information on e.g. group composition could be used for this, and what kind of auxiliary data would need to be collected to reliably model misclassification error.

The second (and perhaps most attractive) aspect of our modelling framework in the context of future work is its spatio-temporal hierarchical structure. While we included spatial, temporal, and residual variation in our framework here, we treated them as independent. Alternatively, spatial (and temporal) correlations among parameters can be modelled explicitly, something that is commonly done e.g. for modern species distribution models [e.g. @pacifici2017; @guelat2018]. For demographic models, this has rarely been implemented so far, not least due to the fact that few demographic models have sufficient spatial resolution [@schaub2021]. The ptarmigan IDSM presented in this study, however, does have sufficient resolution and our results do indeed support that there is spatial clustering in both overall and time-dependent demographic parameters (e.g. @fig-pVR, SFs "Avg_pSurv_Map.pdf" & "Avg_rRep_Map.pdf"). Furthermore, we did find that mixing of several of the global and area-specific intercept parameters in the current model was suboptimal, suggesting that there may be much to gain from additional structuring, as well as from development of more efficient MCMC sampling strategies for the resulting extended model. One promising framework for approaching this are conditionally autoregressive models (CARs). Such models have been used repeatedly for modelling spatial autocorrelation in species occupancy and demographic rates [e.g. @saracco2010; @saracco2012; @guelat2018] and are straightforward to implement using NIMBLE [@lawson2020]. One possible challenge with using CAR models to explicitly model spatial correlations within our ptarmigan IDSM is that CAR models rely on "neighborhood" relationships between discrete areas and many "neighbors" are missing in our ptarmigan data (e.g. @fig-dens). While estimation of latent parameters in missing areas may possible (Perry de Valpine, personal communication; @schaub2021 chapter 19), this also provides an opportunity for inclusion of additional data. The line transect survey data included in this study constitutes just the publicly available part of the data collected through "Hønsefuglportalen" but the programme also includes additional surveys on private land. \[...\] <!-- @ERLEND: Can we add a very brief ocverview over what the coverage (in terms of municipalities) would be if we had access to all the Hønsefuglportalen data? --> Exploring to what degree additional data from Hønsefuglportalen could be included in future studies employing an extended IDSM with additional spatial structuring is therefore a worthwhile endeavor.

Finally, including further data beyond the line transect surveys may be relevant in the future, and in particular in the context of informing and improving management of ptarmigan hunting. In the present study, we have used auxiliary radio-telemetry data to supplement information on survival, but since this data was available for only one out of 41 areas, its influence was likely small. Nonetheless, this illustrates a way for how smaller datasets from single or subsets of areas can be integrated into a large-scale modelling framework. Other relevant data could be included using the same approach [...] <!-- @ERLEND: Any thoughts on what the status here is? Might there be other data available from Lierne and/or beyond (other telemetry data, nest monitoring, etc.? Are there exciting future prospects e.g. linked to automated monitoring with cameras/sound boxes at a larger scale? --> The most relevant source of data to be included into the IDSM framework in the near future, however, is harvest data. <!-- @ERLEND 1-2 sentence summary of what kind of harvest data is available, and at what spatial covarage and resolution --> As the IDSM framework is, in essence, an IPM, harvest can be modelled through partitioning of survival into cause-specific mortality in the process model and inclusion of relevant harvest data likelihoods [e.g. @gamelon2021; @nater2021]. Harvest effects on willow ptarmigan have been studied previously yet uncertainties persist [@sandercock_etal2011; @aanes2002; @pedersen2004; @willebrand2001]. For example, the strength and shape of density dependence and how harvest may induce population sinks because of Allee effects [@kvasnes2015]). Modelling allee effects, or better yet predicting where they are likely to occur if over-harvest occurs, requires a spatially-explicit understanding of harvest and its impact on local populations. Because harvest effort, thus rates, can vary dramatically over time and space [@breisjøberget2018] easily repeatable analyses that can be done continually would prevent prolonged over-exploitation. Harvest effects may interact with other (emergent) factors such as climate change and habitat degradation making the need to be identify triggers for policy changes paramount [@gamelon2019].

<!-- @JAMES & Erlend: Maybe we can be a bit more specific about what the "uncertainties" are regarding ptarmigan harvest? Might be good to keep the "spatially explicit" perspective a bit, i.e. why would it be great to look at this with this kind of model, instead of just aggregate numbers across large regions? Also, maybe we should highlight that up-to-date (easilty repeatable) analyses are important because things can change, and harvest effects may interact with other (emergent) factors such as climate change and habitat degradation. Finally, it could also be nice to link this to some studies on other species... -->

<!--# That do it? -->

@@

Building on the work of @nilsen2023, we applied a novel integrated population model to data collected through a national-scale participatory monitoring programme to estimate spatial and temporal variation in demography of a culturally important game bird species, the willow ptarmigan. While our study was exploratory in nature, it recovered patterns consistent with ecological and life-history theory including trade-offs between survival and recruitment, and a tendency towards slower life histories at higher latitudes and altitudes. Space-for-time substitution also provided the statistical power necessary for the analysis to provide evidence for the alternative prey hypothesis, i.e. ptarmigans benefiting from high abundance of alternative prey for their predators.

<!-- To me, this little paragraph has "abstract vibes", or maybe "conclusion vibes". I'll leave it here for now but note that it may get moved eventually. -->
